My title

### 2024-11-14 paper/paper.qmd
fix reference errors
first draft of paper update

---
title: "A Poll-of-Polls Forecast for the 2024 U.S. Presidential Election: Analyzing Support Trends for Harris by Pollster and State"
subtitle: "Kamala Harris Leads with Stable Support Around 48%"
author: 
  - Yunkyung Ko
thanks: "Code and data are available at: [https://github.com/koyunkyung/us_election_2024](https://github.com/koyunkyung/us_election_2024)."
date: today
date-format: long
abstract: "In the anticipation of the 2024 U.S. presidential election, this paper analyzes trends in Harris's support, focusing on pollster and state-specific factors. Harris’s support has remained steady around 48%, though key pollsters like Siena/NYT and battleground states such as Michigan show notable deviations. Our goal is to propose a reliable electoral prediction by examining correlations among date, pollster, and state variables and developing predictions through linear and Bayesian models."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
toc-title: "Table of Contents"
toc-depth: 2
toc-location: left
---

@@

library(tidyverse)
library(readr)
library(janitor)
library(lubridate)
library(broom)
library(broom.mixed)
library(modelsummary)
library(rstanarm)
library(splines)
library(ggplot2)
library(knitr)
library(kableExtra)
library(gridExtra)
library(here)
library(usmap)  # For the map of US states
library(viridis)  # For a color palette
library(cowplot)

raw_elections_data <- read_csv(here::here("data/01-raw_data/raw_elections_data.csv"))
harris_data <- read_csv(here::here("data/02-analysis_data/harris_elections_data.csv"))
```

@@

The 2024 U.S. presidential election receives a lot of international attention due to its far-reaching implications, affecting global economy, foreign policy, and social issues like climate change [@euromonitor2024]. This paper aims to predict possible outcomes of the election by analyzing Kamala Harris's support in the polling data with a linear regression and a Bayesian approach.

Our estimand is Harris's support rate in polls, tracked over time and adjusted for pollster and state specific variations. Using a "pooling the polls" approach, we aim to correct for variations across different voter bases [@poolingpolls]. Initial linear models show Harris's support rate remains stable at around 48% over time, though with large variability as seen in the spread of points around the fitted line. Some pollsters or states such as Siena/NYT and Michigan consistently reported higher or lower support for Harris compared to others, proposing the importance of accounting for specific polling contexts.

Beyond election forecasting, this analysis signals global stakeholders on potential shifts in the U.S. policy and strategy [@csis2024]. As such, this study not only contributes to the domestic political discourse but also provides a valuable tool for global actors seeking to navigate the uncertainty surrounding the 2024 U.S. presidential election [@csis2024]. The paper is organized as follows: @sec-data and @sec-model detail the data and methodology used, including filtering and modeling techniques applied to the polling data. @sec-results presents findings from the linear and Bayesian models, while @sec-discussion interprets broader implications. Finally, @sec-appendix covers pollster methodology and ideal survey practices.

@@

We use the statistical programming language R [@citeR] to analyze US presidential polling data from FiveThirtyEight [@fivethirtyeight2024], focusing on support for Kamala Harris. The dataset, last updated on 30 October 2024, includes a wide range of poll results from various national and state-level polls, with key variables such as pollster, sample size, percentage of support for Harris, and end date of the poll. Following the guidance of @tellingstories, we compiled the results of each opinion poll over a period of time and compared them taking into account the methodological peculiarities of polling by pollsters and geographical scope of the conducted polls.

To ensure data quality, we filtered the dataset to include only polls that measured Kamala Harris' support, with a numeric grade of pollster 2.7 or higher for reliability. We also limited the analysis to polls conducted after July 21, 2024, when Harris officially declared her candidacy, and excluded pollsters with fewer than 30 polls to focus on those with sufficient data for robust results.

In performing the analysis, we utilized several R packages. `tidyverse` [@tidyverse] was used for data manipulation and visualization and `rstanarm` [@rstanarm], `modelsummary` [@modelsummary] was respectively used for Bayesian modeling and generating model summaries. For visualizing results, `ggplot2` [@ggplot2] was used and `kableExtra` [@kableExtra] helped format tables for presentation. These packages provided a framework for efficient data processing, modeling, and reporting.

@@

The original dataset sourced from FiveThirtyEight [@fivethirtyeight2024] aggregates a wide range of poll results, showing 16817 observations based on dataset available at October 30. The polls conducted by various polling organizations capture voter preferences by taking a representative sample of the electorate and asking for the voters' candidate of choice. Surveys were conducted at the state and national levels, providing the wide perspective on public feelings across the country.

Each poll represents a predictor of an actual event, namely voter opinion at a particular moment. Nevertheless, like all survey results, the raw data is susceptible to many potential limitations including the following: sampling error, variation in polling methods, distortion because of inappropriate survey responses such as missing data or response from respondents who misunderstood the questions [@tellingstories].

While applying several filters to the original dataset such as restricting to those with a numeric grade of 2.7 or higher or pollsters with more than 30 polls improves data reliability, certain limitations still exist. Selection bias and sampling error remains as a concern, since polls always represent only part of the population. Differences in the way different organizations conducted their polling might introduce more inconsistencies. Finally, by focusing our attention on post-declaration polls only, we exclude earlier trends that could add more insight into how Harris' support has evolved over time.

## Outcome variable

### The Proportion (%) of Support that a Candidate Received in the Poll

The main variable of interest that we aim to forecast is the 'pct' variable, which represents the proportion of vote or support that a candidate received in the poll. @tbl-pctraw1 and @fig-pct2 respectively shows the summary statistics and distribution of the 'pct' variable in the original dataset [@fivethirtyeight2024]. @tbl-pctfilter1 and @fig-pctfilter2 shows the summary statistics and distribution of the same variable, but in the filtered dataset that only comprises of the supporting votes for Harris from relatively high-quality polling organizations. Comparing the summary statistics for the raw data (@tbl-pctraw1) and filtered data (@tbl-pctfilter1), higher numbers were derived from data filtered only by Harris supporters. Also, @fig-pctfilter2 illustrates that a significant number of polls indicate support levels ranging from 45% to 50%, which suggests a stable yet not substantial endorsement. This proposes that Harris possesses a reliable foundational support, although her capacity to obtain a majority remains ambiguous.

```{r}
#| label: tbl-pctraw1
#| tbl-cap: Summary statistics for the proportion (%) of support that candidates received in the poll
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

# Calculate summary statistics for the 'pct' variable
summary_stats <- raw_elections_data %>%
  summarize(
    mean = round(mean(pct, na.rm = TRUE), 2),
    median = round(median(pct, na.rm = TRUE),2),
    min = round(min(pct, na.rm = TRUE), 2),
    max = round(max(pct, na.rm = TRUE), 2),
    sd = round(sd(pct, na.rm = TRUE), 2),
    n = n()
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```
```{r}
#| label: fig-pct2
#| fig-cap: Distribution of the proportion (%) of support that candidates received in the poll
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Plot the distribution of 'pct' using a histogram
ggplot(raw_elections_data, aes(x = pct)) +
  geom_histogram(binwidth = 3, fill = "blue", alpha = 0.7, color = "black") +
  theme_minimal() +
  labs(x = "Percentage of Support (pct)",
       y = "Count")

```
```{r}
#| label: tbl-pctfilter1
#| tbl-cap: Summary statistics for the proportion (%) of support that Harris received in high-quality polls
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

# Calculate summary statistics for the 'pct' variable
summary_stats <- harris_data %>%
  summarize(
    mean = round(mean(pct, na.rm = TRUE), 2),
    median = round(median(pct, na.rm = TRUE), 2),
    min = round(min(pct, na.rm = TRUE), 2),
    max = round(max(pct, na.rm = TRUE), 2),
    sd = round(sd(pct, na.rm = TRUE), 2),
    n = n()
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```{r}
#| label: fig-pctfilter2
#| fig-cap: Distribution of the proportion (%) of support that Harris received in high-quality polls
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Plot the distribution of 'pct' using a histogram
ggplot(harris_data, aes(x = pct)) +
  geom_histogram(binwidth = 1.5, fill = "blue", alpha = 0.7, color = "black") +
  theme_minimal() +
  labs(x = "Percentage of Support (pct)",
       y = "Count")

@@

### The Date the Poll was Concluded

The 'end_date' variable representing the time the poll was concluded was put into account to keep track of how support for a candidate changes in time. The reported end dates in the original dataset [@fivethirtyeight2024] ranges from 7 April, 2021 to 29 October, 2024 (@tbl-date). @fig-date shows that the polling data is more concentrated on survey results conducted in the recent period.

```{r}
#| label: tbl-date
#| tbl-cap: Summary statistics for the date that the poll was concluded
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

# Convert 'end_date' to a proper date format
raw_elections_dataa <- raw_elections_data %>%
  mutate(end_date = mdy(end_date))
filtered_data <- raw_elections_dataa %>%
  filter(end_date >= as.Date("2023-01-01"))

summary_stats <- raw_elections_dataa %>%
  summarize(
    Min = min(end_date, na.rm = TRUE),
    Max = max(end_date, na.rm = TRUE)
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

@@

```{r}
#| label: fig-date
#| fig-cap: Distribution of the date that the poll was concluded
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Distribution of 'end_date'
ggplot(filtered_data, aes(x = end_date)) +
  geom_histogram(fill = "grey", alpha = 0.7, color = "black", binwidth = 30) +  # Set binwidth explicitly
  theme_minimal() +
  labs(x = "End Date",
       y = "Count")

When filtering the data, not only pollster quality and candidate type but also the date variable was considered. The filtered data contains only the polling data after the declaration of Harris. So, the date variable for filtered data in @tbl-date2 ranges from 23 July, 2024 to 29 October, 2024. @fig-date2 shows that overall, polling is conducted regularly but intensifies around specific dates.

```{r}
#| label: tbl-date2
#| tbl-cap: Summary statistics for the date that high-quality polls for Harris was concluded
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

summary_stats <- harris_data %>%
  summarize(
    Min = min(end_date, na.rm = TRUE),
    Max = max(end_date, na.rm = TRUE)
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```{r}
#| label: fig-date2
#| fig-cap: Distribution of the date that high-quality polls for Harris was concluded
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Convert 'end_date' to a proper date format
harris_dataa <- harris_data %>%
  mutate(end_date = ymd(end_date))  # Change mdy() to ymd() or dmy() depending on the format

# Distribution of 'end_date'
ggplot(harris_dataa, aes(x = end_date)) +
  geom_histogram(fill = "grey", alpha = 0.7, color = "black", binwidth = 10) +  # Set binwidth explicitly
  theme_minimal() +
  labs(x = "End Date",
       y = "Count")

### Pollster and State

The 'pollster' and 'state' variable were selected to consider the effect of changes in poll-making organizations and geographical distinctions. The two variables respectively represent the polling organization that conducted the poll and the US state where the poll was conducted or focused.

@tbl-predictors1 shows that the original dataset [@fivethirtyeight2024] contains 232 distinct pollsters and 56 distinct states. After filtering for high-quality polls and assigning 'other' for states with fewer than 60 polls, the analysis data contains 8 distinct poll-making organizations and 20 geographical distinctions as shown in @tbl-predictors2.

```{r}
#| label: tbl-predictors1
#| tbl-cap: Number of distinct polling organizations and US states where the poll was conducted
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

summary_stats <- raw_elections_data %>%
  summarize(
    Pollster = n_distinct(pollster),
    State = n_distinct(state)
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```{r}
#| label: tbl-predictors2
#| tbl-cap: Number of distinct high-quality polling organizations and US states where more than 60 polls for Harris were conducted
#| echo: false
#| tbl-align: center
#| tbl-width: 80%

summary_stats <- harris_data %>%
  summarize(
    Pollster = n_distinct(pollster),
    State = n_distinct(state)
  )

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

The distribution of polling counts for different pollsters in @fig-pollster suggests that the analysis data is dominated by a few pollsters, particularly Siena/NYT. Depending on their polling methodology, the general results may have potential biases. A detailed analysis of the polling methodology and possible errors of the organization will be covered in @sec-appendix.

```{r}
#| label: fig-pollster
#| fig-cap: Distribution of polling organizations where high-quality polls for Harris were conducted
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Distribution of 'pollster' (bar plot since it's categorical)
ggplot(harris_data, aes(x = pollster)) +
  geom_bar(fill = "purple", alpha = 0.7, color = "black") +
  theme_minimal() +
  labs(x = "Pollster",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

@fig-state displays the distribution of polls across different states in the analysis data. Pennsylvania, Arizona, and Georgia are the top 3 states with high number of polls while states like Montana, New Mexico, and Maryland have much fewer polls. Note that a significant number of national or unspecified state-level polls are aggregated in this analysis data regarding the high count in 'Other' category. The concentration of polls in certain states further suggests a strategic focus on areas likely to impact the election outcome [@11alive2024].

```{r}
#| label: fig-state
#| fig-cap: Distribution of US states where the more than 60 polls for Harris were conducted
#| echo: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Distribution of 'state' (bar plot)
ggplot(harris_data, aes(x = reorder(state, -table(state)[state]))) +
  geom_bar(fill = "black", alpha = 0.7, color = "black") +
  theme_minimal() +
  labs(x = "State",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

## Correlation between predictor variables

By examining the relationships between predictor variables in the analysis dataset, we aim to identify potential biases in interpretation and model instability. This is expected to ensure a transparent understanding of the model's robustness and interpretability.

### End Date and State

@fig-correlation1 shows notable polling concentrations on certain states such as Pennsylvania, Ohio, and Nevada in week 2024-37 (September 9, 2024) or 2024-38 (September 15, 2024). This suggests that these states are key battelgrounds or areas of strategic focus during the election period. Moreover, polling activities are not consistent across all weeks and there are clear peaks in polling activity in week 2024-38 (September 15, 2024), which may correspond to significant political events, debates, or media focuses.

```{r}
#| label: fig-correlation1
#| fig-cap: "Polling concentration over time by state (*Note:* Strategic focus of polling is directed in battleground states such as Pennsylvania and Ohio in September 15, 2024)"
#| echo: false
#| fig-align: center
#| page-break-after: true

# Aggregate the number of polls by state and by week (or month)
poll_time_state_2 <- harris_data %>%
  mutate(week = format(as.Date(end_date), "%Y-%U")) %>%  # Group by week of the year
  group_by(state, week) %>%
  summarise(total_polls = n(), .groups = "drop")  # Adding .groups = "drop"

# Plot a heatmap of polling activity by state and week
ggplot(poll_time_state_2, aes(x = week, y = state, fill = total_polls)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "grey90") +
  labs(x = "Week",
       y = "State",
       fill = "Total Polls") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 8))

### End Date and Pollster

@fig-correlation2 shows that Siena/NYT, which made up the largest proportion of polls in our analysis dataset, has concentrated polling efforts in week 2024-37 (September 9, 2024) and week 2024-38 (September 15, 2024). Notably active periods, possibly around key election events, might introduce temporal bias and overemphasize methodologies of Siena/NYT in that specific time period. Emerson shows overall consistency of polling activities across several weeks indicating steady involvement, while other polling organizations show more sporadic or minimal involvement.

```{r}
#| label: fig-correlation2
#| fig-cap: "Polling concentration over time by pollster (*Note:* Concentrated polling efforts of Siena/NYT are shown in the first half of September, 2024.)"
#| echo: false
#| fig-align: center
#| page-break-after: true

# Aggregate the number of polls by state and by week (or month)
poll_time_pollster <- harris_data %>%
  mutate(week = format(as.Date(end_date), "%Y-%U")) %>%  # Group by week of the year
  group_by(pollster, week) %>%
  summarise(total_polls = n(), .groups = "drop")  # Adding .groups = "drop"

# Plot a heatmap of polling activity by state and week
ggplot(poll_time_pollster, aes(x = week, y = pollster, fill = total_polls)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "grey90") +
  labs(x = "Week",
       y = "Pollster",
       fill = "Total Polls") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 8))

### Pollster and State

@fig-correlation3 shows that Siena/NYT has the strongest state-level presence in key battleground states like Michigan and Arizona. Others like AtlasIntel and Beacon/Shaw has minimal polling coverage in targeted states such as Nevada and Georgia. This variation in coverage could influence the overall analysis in that some states might receive disproportionately more attention from certain pollsters.

```{r}
#| label: fig-correlation3
#| fig-cap: "Polling activity by pollster and state (*Note:* Siena/NYT has strongest state-level polling coverage in key battleground states)"
#| echo: false
#| page-break-after: true

# Step 1: Filter out NAs from pollster column and summarize data by state and pollster
pollster_state_data <- harris_data %>%
  filter(!is.na(pollster)) %>%  # Remove NAs from pollster
  group_by(state, pollster) %>%
  summarise(total_polls = n(), .groups = "drop")  # Summarize number of polls

# Step 2: Plot the faceted map using plot_usmap
plot_usmap(data = pollster_state_data, values = "total_polls", regions = "states") +
  facet_wrap(~ pollster, ncol = 4) +  # Create a map for each pollster
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "gray90") +  # Use blue color palette
  labs(fill = "Total Polls") +
  theme_void() +  # Remove axes and gridlines
  theme(strip.text = element_text(size = 9))  # Customize facet labels

# Model {#sec-model}

Our modelling strategy estimates Kamala Harris's support percentage in the 2024 US election polls, accounting for potential variations over time including pollster and state differences. The model balances complexity with interpretability, using both linear and Bayesian frameworks to capture patterns in the data. We turned from the initial linear model to a Bayesian model for robust predictions considering the variables' uncertainties. The Bayesian approach, which incorporates both prior beliefs and observed data to make predictions, allowed us to capture the variances among pollsters and states more accurately and responsively. Further background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the percentage of support that Kamala Harris receives in poll $i$. We begin with two simple linear models and progress to more complex Bayesian models that account for hierarchical structures.

The following models outline our approach:

### Linear Model by Date

```{=latex}
\begin{align}  
y_i &= \beta_0 + \beta_1 \cdot \text{end\_date}_i + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}
```

where:
- $y_i$ is the percentage of support for Harris in poll $i$,
- $\beta_0$ is the intercept,
- $\beta_1$ represents the effect of the poll's end date,
- $\epsilon_i$ is the error term.

### Linear Model by Date and Pollster

```{=latex}
\begin{align}  
y_i &= \beta_0 + \beta_1 \cdot \text{end\_date}_i + \gamma_{p[i]} + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}
```

where:
- $\gamma_{p[i]}$ is a fixed effect for pollster $p$ conducting poll $i$ (e.g., Siena/NYT).

### Bayesian Model with Random Intercept for Pollster

```{=latex}
\begin{align}  
y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \cdot \text{end\_date}_i + \gamma_{p[i]} \\
\gamma_p &\sim \text{Normal}(0, \sigma_{\gamma})
\end{align}
```

where:
- $\gamma_p$ is a random effect for pollster $p$.

### Bayesian Model with Random Intercept for Pollster and State

```{=latex}
\begin{align}  
y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \cdot \text{end\_date}_i + \gamma_{p[i]} + \delta_{s[i]} \\
\gamma_p &\sim \text{Normal}(0, \sigma_{\gamma}) \\
\delta_s &\sim \text{Normal}(0, \sigma_{\delta})
\end{align}
```

where:
- $\delta_{s[i]}$ is a random effect for state $s$.

The Bayesian models are fit using `rstanarm` in R. The priors used are weakly informative:
- $\beta_0 \sim \text{Normal}(0, 10)$
- $\sigma \sim \text{Exponential}(1)$

### Model justification

Different pollsters and states induce variations in polling results, as pollsters may have distinct methodologies and states represent diverse voter bases. Incorporating random effects for both pollsters and states allows us to improve the robustness of the model.

These models are run through the rstanarm package [@rstanarm] in R  [@citeR], which makes Bayesian modeling available through the use of Stan's strong inference engine. To validate the models, RMSE and WAIC have been made use of to check the goodness of fit; Bayesian models with reduced RMSE and WAIC outperform linear models. We use weakly informative priors; for example, $\beta_0 \sim \text{Normal}(0, 10)$ and $\sigma \sim \text{Exponential}(1)$. This reflects our initial uncertainty but prevents overfitting. The priors were chosen conservatively to ensure that the model remains consistent.

Model diagnostics, including posterior predictive checks and convergence diagnostics, were carried out to ensure the reliability of the results. The Bayesian models converged successfully, as indicated by $\hat{R} = 1$ for all parameters.

The main assumption in these models is that the pollster and state effects can be treated as random. This assumes that the effects are normally distributed across pollsters and states, which may not always be accurate. Additionally, the model assumes that polling data is representative of the actual electorate, an assumption that can be violated if polls are biased or have non-random sampling issues. Despite these limitations, the hierarchical structure allows us to capture important variability, making the model suitable for predicting Harris's support. Future improvements could involve incorporating time-varying effects or exploring interactions between pollsters and states.

# Results {#sec-results}

## Results from examining the analysis data

@fig-resultplot1 shows an initial increase in support for Harris since her declaration to join the electoral race, peaking around mid-September. However, this initial momentum stabilized with minor decrease in support as the election gets closer. Moreover, the scattered data points around the summary line show high variance between individual polls.

```{r}
#| label: fig-resultplot1
#| fig-cap: "Polling votes for Harris over time (*Note:* The date starts from after Harris officially announced her campaign for 2024 U.S. presidential election on July 21, 2024.)"
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

base_plot <- ggplot(harris_data, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date")

# Plots poll estimates and overall smoothing
base_plot +
  geom_point(alpha = 0.45) +
  geom_smooth()
```

When accounting for the pollster difference, @fig-resultplot2 and @fig-resultplot3 show variations in Harris' support levels that are obscured in the aggregated view of @fig-resultplot1. Siena/NYT (pink) shows relatively stable polling results for Harris, with fewer extreme values and a slight downward trend in October. On the other hand, Emerson (green) shows more fluctuation throughout the period with a wider spread of polling values.

```{r}
#| label: fig-resultplot2
#| fig-cap: "Polling votes for Harris over time by pollster (*Note:* The date starts from after Harris officially announced her campaign for 2024 U.S. presidential election on July 21, 2024. Only filtered pollsters with high-quality polling questions are shown in the following.)"
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Color by pollster
base_plot +
  geom_point(aes(color = pollster), alpha = 0.45) +
  geom_smooth() + 
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-resultplot3
#| fig-cap: "Polling votes for Harris over time by pollster (facets) (*Note:* The date starts from after Harris officially announced her campaign for 2024 U.S. presidential election on July 21, 2024. Only filtered pollsters with high-quality polling questions are shown in the following.)"
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| page-break-after: true

# Facet by pollster
base_plot +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x)) +  # Use GAM smoothing to handle complex data
  facet_wrap(vars(pollster))
```

>>>>

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
```

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

We use the statistical programming language SQL [@citeR]

.... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

### Price Levels

```{r}
#| label: fig-price
#| fig-cap: Price level distributions across all observations used in the analysis
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
apple <- read.csv("../data/analysis_data/avg_price_by_vendor_apple.csv")
mango <- read.csv("../data/analysis_data/avg_price_by_vendor_mango.csv")
pear <- read.csv("../data/analysis_data/avg_price_by_vendor_pear.csv")
orange <- read.csv("../data/analysis_data/avg_price_by_vendor_orange.csv")

vendor_avg_apple <- aggregate(avg_price ~ vendor, data = apple, mean)
vendor_avg_mango <- aggregate(avg_price ~ vendor, data = mango, mean)
vendor_avg_pear <- aggregate(avg_price ~ vendor, data = pear, mean)
vendor_avg_orange <- aggregate(avg_price ~ vendor, data = orange, mean)

vendor_avg_apple$fruit <- "apple"
vendor_avg_mango$fruit <- "mango"
vendor_avg_pear$fruit <- "pear"
vendor_avg_orange$fruit <- "orange"

dataset <- rbind(vendor_avg_apple, vendor_avg_mango, vendor_avg_pear, vendor_avg_orange)
datao <- rbind(apple, mango, pear, orange)

# Plot the distribution of prices with a histogram
ggplot(datao, aes(x = avg_price)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") + # Set bin width and colors
  labs(
       x = "Average Price",
       y = "Frequency") +
  theme_minimal()

@@

### Type of Fruits: The Product that We Specified to Compare the Price Distribution

@fig-fruits shows the frequency of observations for four fruit types: apples, mangoes, oranges, and pears. Apples have the highest count, with around 50 instances, possibly due to popularity or market demand. Pears follow with a moderate count of about 20. Mangoes and oranges have the lowest counts, around 10 each, possibly due to factors like seasonality or regional availability.

```{r}
#| label: fig-fruits
#| fig-cap: The distrubution of different categories of fruits used in the analysis
#| echo: false
#| warning: false
#| message: false

# Count the number of observations in each dataset
counts <- data.frame(
  fruit = c("apple", "mango", "pear", "orange"),
  observation_count = c(nrow(apple), nrow(mango), nrow(pear), nrow(orange))
)

# Plot the observation count for each fruit
ggplot(counts, aes(x = fruit, y = observation_count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
       x = "Fruit Type",
       y = "Observation Count") +
  theme_minimal()

@@

### Type of Vendors

@fig-vendors provides a visualization of the number of observations for each vendor in the dataset, specifically Loblaws, NoFrills, and Voila. The x-axis represents these three vendors, while the y-axis shows the count of observations, likely indicating the number of unique products or price points recorded for each vendor. Each bar is labeled with the exact count: Loblaws has 33 observations, NoFrills has 31, and Voila has the highest at 36. These counts are relatively similar across vendors, suggesting a fairly balanced representation in the data, although Voila has a slight edge in the number of entries. This balanced distribution is beneficial for comparative analysis, as it reduces the risk of one vendor disproportionately influencing results. The slight difference in counts could introduce a minor weight toward Voila in any statistical analysis, but overall, the dataset’s even distribution supports reliable comparisons. Thus, any averages or other metrics calculated by vendor are likely to be comparable across Loblaws, NoFrills, and Voila, minimizing selection bias due to unequal observation counts.

```{r}
#| label: fig-vendors
#| fig-cap: The distribution of vendors used in the analysis
#| echo: false
#| warning: false

# Plot the distribution of observation count by vendor with adjusted bar width and labels
ggplot(datao, aes(x = vendor)) +
  geom_bar(fill = "purple", width = 0.4) + # Narrower bar width
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) + # Add count labels above bars
  labs(
       x = "Vendor",
       y = "Observation Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for readability
```

# Results

@fig-results illustrates the average prices of four fruits—apples, mangoes, oranges, and pears—across three different vendors: Loblaws, NoFrills, and Voila. Each fruit is represented by a distinct color, as indicated in the legend: red for apples, green for mangoes, blue for oranges, and purple for pears. The bar plot component of the chart displays the average price of each fruit at the respective vendors. Notably, apples are priced highest at Loblaws and lowest at Voila. Mangoes exhibit minimal price variation across the vendors. Oranges show a significant price difference, being most expensive at Loblaws and least expensive at Voila. Similarly, pears are priced lowest at Voila. Superimposed on the bar plot is a line plot that connects the average prices of each fruit across the vendors, highlighting the price trends. The lines for apples, oranges, and pears demonstrate a downward trend from Loblaws to Voila, indicating decreasing prices. In contrast, the line for mangoes remains relatively flat, suggesting consistent pricing among the vendors. In summary, the chart effectively compares the pricing strategies of different vendors for various fruits, revealing that Voila generally offers lower prices for most fruits, while Loblaws tends to have higher prices. These visual trends provide insights into the pricing correlations among the vendors and set the stage for further analysis of potential causal factors influencing these pricing differences.

```{r}
#| label: fig-results
#| fig-cap: Price distribution of different kinds of fruits(apple, mango, orange, pear) by vendors
#| echo: false
#| warning: false
#| message: false

ggplot(dataset, aes(x = vendor, y = avg_price, fill = fruit, group = fruit)) +
  geom_bar(stat = "identity", position = "dodge", aes(color = fruit)) +
  geom_line(aes(color = fruit)) +
  geom_point(aes(color = fruit)) +
  facet_wrap(~fruit, scales = "free_y") +
  labs(x = "Render", y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Discussion

## Correlation vs. Causation

When analyzing the pricing of the same fruit in different supermarkets, it is very important to distinguish Correlation and Causation for understanding the root cause of price difference. We have observed that the prices of Loblaws apples, pears, orange and mango are always higher than those of other supermarkets (such as NoFrills and Voila), and the price of each fruit in Voila is always lower than that in other stores, which shows that there is a correlation between supermarket and fruit pricing. This correlation may be related to the market positioning, operating costs and supply chain management of supermarkets. For example, Loblaws is positioned in the high-end market and can provide better products and services to middle-class customers, so its pricing is higher. Voila mainly serves groups with limited budgets, so the price is lower. In addition, factors such as geographical location, rent and employee salary may lead to differences in operating costs, which in turn may affect pricing strategies. For example, Loblaws is located in the city center, the rent is high, and the salary of employees is relatively high because of the high service level. In supply chain management, for example, the procurement channels of different supermarkets and the efficiency of supply chain may lead to cost differences and affect the final selling price. In addition, to determine whether these factors really lead to price differences, further data and analysis are needed, such as market positioning research, cost structure analysis and supply chain evaluation. Through these in-depth analysis, we can understand the causal relationship of price difference more accurately.

## Missing Data

In this dataset, missing data occurs because different supermarkets offer varying selections of fruit varieties. For example, some supermarkets may carry a wide range of apple varieties, while others only stock a few. As a result, when analyzing data across all supermarkets, information on certain apple varieties may be incomplete, with specific types available only from certain vendors. Similar patterns may apply to other types of fruit, leading to gaps in the dataset. This missing data reflects differences in product offerings across vendors, and it's essential to be aware of and consider this incompleteness when conducting further analysis.

## Sources of Bias
In this dataset, we study the price differences and potential changes of 4 fruits between different supermarkets. The price differences are due to the different positioning of each supermarket, Loblaw, NoFrills, and Voila.

### Selection Bias
As mentioned before, selection bias in this dataset mainly comes from the differences in customer groups of different supermarkets. Each supermarket (such as Loblaws, NoFrills, and Voila) has a different market positioning and customer group. Customers who choose to shop at Loblaws usually prefer high-end goods or organic foods, while NoFrills customers may pay more attention to low prices and discounts. This customer preference leads to bias in the price data, that is, the price differences between different supermarkets not only reflect the market price of the product itself, but also are affected by customer selection behavior. Therefore, Loblaws' higher prices do not necessarily mean that all its fruits are priced higher than other supermarkets, but rather reflect that its customers are willing to pay a premium for higher quality or brand.

### Measurement Bias
In this dataset, measurement bias may come from the different ways in which supermarkets record or define the average price (avg_price). For example, Loblaws may be promoting a specific brand or organic fruit, making the recorded average price lower than the actual price, while NoFrills may use a specific price for a single batch, ignoring other higher price ranges. These differences in measurement may cause distortions in price data, making our price comparisons between different supermarkets biased.

### Confounding Bias
In this dataset, confounding bias may come from factors such as fruit type, quality, packaging type, seasonal changes, etc. For example, Loblaws may be more inclined to sell high-quality or organic apples, which are usually priced higher, and this tendency is not shown separately in the data. Therefore, the price difference between supermarkets may not be due to the supermarket itself, but to uncontrolled fruit quality or seasonal factors.
